---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=FALSE}
library(plyr)
library(lattice)
options(scipen=999)
```

Load the data in variable "data" and convert the dates.
```{r}
data = read.csv("activity.csv")
data = mutate(data, date=as.Date(date))
```

## What is mean total number of steps taken per day?
```{r}
stepsperday = ddply(data, .(date), summarize, totalsteps=sum(steps))
hist(stepsperday$totalsteps, 100, main="Histogram of total number of steps each day", xlab="Total steps per day")
```

The mean number of steps taken per day is 
`r round(mean(stepsperday$totalsteps, na.rm=TRUE),1)` and the median number of steps each day is `r round(median(stepsperday$totalsteps, na.rm=TRUE),1)`.


## What is the average daily activity pattern?
```{r}
stepsperinterval = ddply(data, .(interval), summarize, averagesteps=mean(steps,na.rm=TRUE))
with(stepsperinterval,plot(interval, averagesteps, type="l", main="Average number of steps per interval", xlab="Interval", ylab="Number of steps"))
```

The maximum value is achieved at:
```{r}
stepsperinterval[which.max(stepsperinterval$averagesteps),]
```

## Imputing missing values
Total number of missing values in the dataset is `r sum(is.na(data$steps))`.

The strategy for imputing missing data is to replace each NA with the average for the respective interval accross all days.

```{r}
data2 = data
for (i in which(is.na(data$steps))) {
    data2[i,"steps"] <- stepsperinterval[which(stepsperinterval$interval==data[i,"interval"]),"averagesteps"]}
```

```{r}
stepsperday2 = ddply(data2, .(date), summarize, totalsteps=sum(steps))
hist(stepsperday2$totalsteps, 100, main="Histogram of total number of steps each day w/o NAs", 
     xlab="Total steps per day")
```

The new mean number of steps taken per day is 
`r round(mean(stepsperday2$totalsteps, na.rm=TRUE),1)` and the new median number of steps each day is `r round(median(stepsperday2$totalsteps, na.rm=TRUE),1)`.


## Are there differences in activity patterns between weekdays and weekends?
```{r}
weekend = function(x) if(x=="Saturday" || x=="Sunday") return("weekend") else return("weekday")
data2 = mutate(data2, day=as.factor(sapply(weekdays(data2$date), weekend)))
stepsperinterval3 = ddply(data2, .(interval,day), summarize, averagesteps=mean(steps,na.rm=TRUE))
xyplot(averagesteps ~ interval | day, data=stepsperinterval3, type="l", xlab="Interval", ylab="Number of steps", layout= c(1,2))

```
